\large{Curtin University} \\ \large{Curtin Institute of Radio Astronomy} \\ \large{Physics Project 1} \\ \vspace*{20px} \textbf{\LARGE{Finding New Pulsars Using Machine Learning}}\vspace*{10px}

### 2020-06-03 Draft.tex
Methods and appendices

\verb|PulsarValidator| is designed to load the positive and negative output files generated from the machine learning classification of a controlled testing dataset of pulsar candidates, and compare these files to a list of the pulsars included in the testing dataset. The software outputs the following statistics: the number of pulsars and non-pulsars, the number of detected pulsars and non-pulsars, and the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).

@@

\subsection{Evaluating the LOTAAS Machine Learning Classification Tool for use with the SMART Survey}

To address the Candidate Selection Problem discussed in \emph{Section 1.3} of this report, a Java Machine Learning pulsar classification tool named \verb|LOTAASClassifier| was created to classify the pulsar candidates produced by the LOFAR Tied-Array All-sky Survey (LOTAAS) \autocite{lyon}. Due to the similarities between the LOFAR radio telescope and the Murchison Widefield Array (MWA) radio telescope, it is logical to attempt to apply the \verb|LOTAASClassifier| tool to candidates produced by the MWA.

The \verb|LOTAASClassifier| tool contains four machine learning algorithms with which the user can choose to make pulsar classifications. They are: the J48 (C4.5 Decision Tree \autocite{quinlan}), the Multi-Layer Perceptron, the Naive Bayes, and the Support Vector Machines (SVM) algorithms \autocite{lyon}.

To begin evaluating the software, \verb|LOTAASClassifier v1.0| can be downloaded from its GitHub repository \\(\verb|https://github.com/scienceguyrob/LOTAASClassifier|) with the following UNIX terminal commands:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars
$ git clone https://github.com/scienceguyrob/LOTAASClassifier.git
\end{lstlisting}

The tool's executable program can then be found by navigating to the following directory in the UNIX terminal:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/LOTAASClassifier/dist
\end{lstlisting}

\subsubsection{Evaluating LOTAASClassifier}

With a successfully created machine learning classification model, it is now possible to test \verb|LOTAASClassifier| against a previously unseen (by the classifier) dataset of Murchison Widefield Array candidates whose class is also known. We can first remove the existing candidates from inside the \verb|PFD| files directory and delete the existing \verb|output| files.

The candidates directory can then be populated by new \verb|PFD| files and we can use the `\verb|docker run|' command from \emph{Section 2.2.1} to extract the features from the new candidates.

The \verb|LOTAASClassifier| tool can now make classification predictions on the new candidates by running the UNIX terminal command:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/LOTAASClassifier/dist
$ java -jar LOTAASClassifier.jar -p ~/pulsars/data/pfd/output.arff -m ~/pulsars/data/model.m -a 1
\end{lstlisting}

This can be repeated for each machine learning classifier present in the \verb|LOTAASClassifier| software by changing the argument \verb|-a| from 1 through to 4.

We can now run the \verb|PulsarValidator| program created in \emph{Section 2.2.1} to evaluate the accuracy of \verb|LOTAASClassifier|'s algorithms by running the command:

\begin{lstlisting}[numbers=none]
$ cd ~/pulsars/PulsarValidator/target
$ java -jar pulsarvalidator-1.0-full.jar -v ~/pulsars/data/pfd/pulsars.txt ~/pulsars/data/pfd/output_[classifier].positive ~/pulsars/data/pfd/output_[classifier].negative
\end{lstlisting}

Where the \verb|[classifier]| variable can be changed depending on the name of the classifier algorithm being evaluated. The outputted validation statistics can then be compared across the algorithms to determine their usefulness with the Murchison Widefield Array.

\subsubsection{Creating the PulsarClassifier Ensemble Classification Tool}
\label{sec:methodensemble}
According to \cite{tan}, using Machine Learning algorithms in ensemble to make pulsar classifications increases the accuracy of classifications, classifying pulsars that were often misclassified such as wide-pulse pulsars.

To build the ensemble classification feature into the existing \verb|LOTAASClassifier| tool, we can first begin by creating a new Java project named \verb|PulsarClassifier| using the free software, Maven (\verb|https://maven.apache.org|).

@@

\pagebreak
\begin{lstlisting}[numbers=none, language=xml, title=pom.xml, basicstyle=\footnotesize\ttfamily]
<project>
    ...
    <build>
      <pluginManagement>
        <plugins>
          ...

The complete source code and build of \verb|PulsarClassifier| can be found at \verb|https://github.com/jacob-ian/PulsarClassifier.git|, or in \emph{Appendix \ref{sec:pulsarclassifier}}.

\subsubsection{Evaluating PulsarClassifier}

Now that we have successfully built \verb|PulsarClassifier|, we can evaluate its usage with the Murchison Widefield Array's (MWA) candidates. We begin by:
\begin{enumerate}[label=\roman*.] 
    \item deleting the previous \verb|output.positive| and \verb|output.negative| files from the \verb|PFD| candidates directory; and
    \item deleting the previous \verb|model.m| file created by the \verb|LOTAASClassifier| software.
\end{enumerate}

\verb|PulsarClassifier| will create the prediction output files for each classifier, and then the \verb|output_ensemble.positive| and \verb|output_ensemble.negative| files for the ensemble classifier. We can validate the ensemble classifier with the \verb|PulsarValidator| program created in \emph{Section \ref{sec:methodvalidate}} by running the commands:

@@

        \subsection{Creating a Classification Model with LOTAASClassifier}

        In order to use \verb|LOTAASClassifier| to classify Murchison Widefield Array candidates, we must use the Machine Learning Training Dataset created in \emph{Section 2.1} to create a classification model. This can be completed by running the following commands:

        \begin{lstlisting}[numbers=none]
        $ java -jar LOTAASClassifier.jar -t ~/pulsars/data/trainingSet.arff -m ~/pulsars/data/model.m -a 1
        \end{lstlisting}

        This instructs \verb|LOTAASClassifier| to create a new classification model for the J48 machine learning algorithm with the previously constructed training dataset.

        To ensure that the classification model was created successfully, we can test the model against the same dataset provided by N. Swainston to create the training dataset. Using \verb|PulsarFeatureLab| we can produce a new \verb|output.arff| file with the candidates' class unedited.

        It is now possible to test the classification model with the candidate feature sets with the following commands:

        \begin{lstlisting}[numbers=none]
        $ cd ~/pulsars/LOTAASClassifier/dist
        $ java -jar LOTAASClassifier.jar -p ~/pulsars/data/pfd/output.arff -m ~/pulsars/data/model.m -a 1
        \end{lstlisting}

        The \verb|LOTAASClassifier| tool will output two files: `\verb|output.positive|' and `\verb|output.negative|', in the same location as the input dataset. The J48 classification model was constructed successfully if the candidates inside the `\verb|positive|' file are the known pulsars.

>>>>

\verb|PulsarValidator| was designed to load the positive and negative output files generated from the machine learning classification of a controlled testing dataset of pulsar candidates, and compare these files to a list of the pulsars included in the testing dataset. The software outputs the following statistics: the number of pulsars and non-pulsars, the number of detected pulsars and non-pulsars, and the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN).

@@

\subsection{Developing an Ensemble Machine Learning Classifier}

To implement the ensemble classifier feature outlined by \textcite{tan} to the \verb|LOTAASClassifier| software, a new Java program named \verb|PulsarClassifier| was developed.

\verb|PulsarClassifier| was designed so that a new `ensemble classifier' option can be chosen as the classification algorithm as well as the previously existing algorithms from \verb|LOTAASClassifier|. The program outputs the classification results of all machine learning classifiers from the \verb|LOTAASClassifier| program, as well as that of the new ensemble classifier. This was achieved by creating the following new Java classes: `PulsarClassifier', `ClassPredictor', `ClassifierBuilder', and `ClassifierValidator', and extending the existing versions of those classes from \verb|LOTAASClassifier| to maintain some of the prior methods and properties.

The Java classes `ClassifierBuilder' and `ClassifierValidator' were modified such that if the user selected the ensemble classifier as the classification algorithm, the method involved in building or validating a machine learning classifier would iterate through the list of available machine learning classifiers (C4.5, Multi-Layered Perceptron, Naïve Bayes, and Support Vector Machine) to build a classification model for each ML classifier, or validate each ML classifier, respectively. These processes can be seen in \emph{Algorithm \ref{alg:classifierbuilder}} and \emph{Algorithm \ref{alg:classifiervalidator}}.

\begin{algorithm}[H]
    \caption{ClassifierBuilder (pseudocode)}
    \label{alg:classifierbuilder}
    \begin{algorithmic}[1]
        \If{$algorithm = -1$}
            \For{each algorithm $i=1\rightarrow 4$}
                \Comment Loop through all classifiers
                \State buildClassifier($i$, $trainingSet$, $modelsDirectory$)
            \EndFor
        \Else 
            \Comment Build individual classifier
            \State buildClassifier($algorithm$, $trainingSet$, $modelPath$)

        \EndIf
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
    \caption{ClassifierValidator (pseudocode)}
    \label{alg:classifiervalidator}
    \begin{algorithmic}[1]
        \If{$algorithm = -1$}
            \For{each algorithm $i=1\rightarrow 4$}
                \Comment Loop through all classifiers
                \State testClassifier($i$, $testSet$, $modelsDirectory$)
            \EndFor
        \Else 
            \Comment Test individual classifier
            \State testClassifier($algorithm$, $testSet$, $modelPath$)
        \EndIf

@@

The Java class `ClassPredictor' was modified such that if the user chooses to use the ensemble classifier, a Java `ArrayList' is created to store the filepaths to the outputs of each machine learning classifier. The list of available machine learning classifiers is then iterated over, generating the classification predictions from each classifier and adding their outputs to the created list of output files. The Java classes `Classification' and `ClassificationList' are then created to be a key-value pair data class and a list of key-value classifications respectively, where the key denotes the filename of a candidate and the value denotes the number of times this candidate has been given a particular classification. A `ClassificationList' for positive and negative classifications are then constructed and the candidates in the positive and negative classifier output files are iterated over, adding the candidates to their respective lists and incrementing their associated value for every recurring classification. As discussed in \textcite{tan}, a positive classification in an ensemble classifier is generally made when a candidate has received a positive classification from 3 separate classifiers. To implement this selection criteria, the positive `ClassificationList' is iterated over and the candidates with an associated value less than 3 are moved into the negative `ClassificationList'. The remaining positive and negative `ClassificationList's are then outputted to their respective positive and negative output files for the ensemble classifier. This process can be seen in \emph{Algorithm \ref{alg:classpredictor}}.

\begin{algorithm}
    \caption{ClassPredictor (pseudocode)}
    \label{alg:classpredictor}
    \begin{algorithmic}[1]
        \If{$algorithm = -1$}
            \State list = new OutputFileList()
            \For{algorithm $i=1\rightarrow4$}
                \Comment Loop through all classifiers
                \State makePredictions($i$, $inputData$, $modelsDirectory$)
                \State list.add(ClassifierOutputFiles)
                \Comment Add output filepaths to list
            \EndFor
            \State positiveList = new ClassificationsList()
            \State negativeList = new ClassificationsList()
            \Comment Create classification lists
            \For{each OutputFile from list}
                \If{OutputFile is .positive}
                    \For{each line in OutputFile}
                        \State positiveList.add(line)
                        \Comment Add to +ve classifications
                    \EndFor
                \ElsIf{OutputFile is .negative}
                    \For{each line in OutputFile}
                        \State negativeList.add(line)
                        \Comment Add to -ve classifications
                    \EndFor
                \EndIf
            \EndFor
            \For{each classification in positiveList}
                \If{classification.occurrences $< 3$}
                    \State negativeList.add(classification)
                    \Comment cut-off at $< 3$ classifications
                \Else
                    \State positiveOutput(classification)
                    \Comment Output classified as Pulsar
                \EndIf
            \EndFor
            \For{each classification in negativeList}
                \State negativeOutput(classification)
                \Comment Output classified as non-Pulsar
            \EndFor

        \Else 
            \State makePredictions($algorithm$)
            \Comment Use individual classifier
        \EndIf
    \end{algorithmic}
\end{algorithm}

The final modification that was made included changing the Java class `PulsarClassifier' to include the ensemble classifier as an option in the user selectable algorithms. Further details on the development of this software can be found in \emph{Appendix \ref{apndx:pulsarclassifier}} and the full Java source code can be found in \emph{Appendix \ref{src:pulsarclassifier}} or on GitHub (\verb|https://github.com/jacob-ian/PulsarClassifier.git|).

\subsection{Evaluating the Machine Learning Classifiers}

A set of 147 pulsar and 87 non-pulsar candidates generated by the SMART survey were then provided by N. Swainston to test the machine learning classifiers on Murchison Widefield Array data. To test all machine learning classifiers present in \verb|LOTAASClassifier|, and the ensemble classifier, the new \\\verb|PulsarClassifier| program was used.

Before making predictions on the test data, it is necessary to create the machine learning classification models from the training dataset. Using \\\verb|PulsarClassifier|, the training dataset was used to generate the classification models for each ML classifier. \verb|PulsarFeatureLab| was then used to extract the machine learning features from the testing dataset of candidates. \verb|PulsarClassifier| was then used to classify the candidates in the testing dataset with the ensemble classifier (and hence with each of the other ML classifiers). For details on the use of \verb|PulsarClassifier|, see \emph{Appendix \ref{apndx:usepulsarclassifier}}.

To evaluate the accuracy of the predictions of the machine learning classifiers, a list of the pulsars included in the testing dataset was created and \verb|PulsarValidator| was ran with the output files of each classifier. For details on the use of \verb|PulsarValidator|, see \emph{Appendix \ref{apndx:usepulsarvalidator}}.

@@

        \subsection{Developing PulsarClassifier}
        \label{apndx:pulsarclassifier}
        To build the ensemble classification feature into the existing \verb|LOTAASClassifier| tool, we can first begin by creating a new Java project named \verb|PulsarClassifier| using the free software, Maven (\verb|https://maven.apache.org|).

\begin{lstlisting}[numbers=none, language=xml, title=pom.xml, basicstyle=\footnotesize\ttfamily]
<project>
    ...
    <build>
      <pluginManagement>
        <plugins>
          ...

The complete source code and build of \verb|PulsarClassifier| can be found at \verb|https://github.com/jacob-ian/PulsarClassifier.git|, or in \emph{Appendix \ref{src:pulsarclassifier}}.

\subsection{Using PulsarClassifier}
\label{apndx:usepulsarclassifier}

We can use \verb|PulsarClassifier| with the Murchison Widefield Array's (MWA) candidates by:
\begin{enumerate}[label=\roman*.] 
    \item deleting the previous \verb|output.positive| and \verb|output.negative| files from the \verb|PFD| candidates directory; and
    \item deleting the previous \verb|model.m| file created by the \verb|LOTAASClassifier| software.
\end{enumerate}

\verb|PulsarClassifier| will create the prediction output files for each classifier, and then the \verb|output_ensemble.positive| and \verb|output_ensemble.negative| files for the ensemble classifier.

\subsection{Using PulsarValidator}
\label{apndx:usepulsarvalidator}
We can validate the ensemble classifier with the \verb|PulsarValidator| program created in \emph{Section \ref{sec:methodvalidate}} by running the commands:
